\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-18 paper.tex
Tweaks from Gregor

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
% New developments have made it possible to infer ARGs at scale,
% enabling many new applications in population and statistical genetics.
% This rapid progress has led to a diversity of ARG definitions and representations.
Classical formalisms have focused on mapping 
coalescence and recombination events to the nodes in an ARG.
We show that this approach is out of step with modern developments,
which do not represent genetic inheritance in terms of these events 
or explicitly infer them.
We present a simple formalism that defines an ARG in terms 
of specific genomes and their intervals of genetic inheritance,
and show how it generalises these classical treatments and 
encompasses the outputs of recent methods. We discuss 
nuances arising from this more general structure, and argue that it
forms an appropriate basis for a software standard in this rapidly 
growing field.
\end{abstract}

@@

\section{Event ARGs}
\label{sec-eARG}
A common, classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This Event ARG (eARG) encoding is the basis of the output formats 
created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG.
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
backwards in time. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
we have three genome segments that are ancestral to the
four samples. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

\section{A diversity of structures}
\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring}
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
The ARGs in Fig.~\ref{fig-inferred-args}A and B are precise 
estimates (Appendix~\ref{sec-precision}),
with each node corresponding to a common ancestor 
or recombination event, or  equivalently, either having two children
or two parents.
In contrast the ARGs in Fig.~\ref{fig-inferred-args}C and D do not 
have this clear-cut interpretation, and the nodes
can simultaneously have more that than two children and more
than two parents. Another dimension of variability among the ARGs
is that the first three methods infer nodes that have
a ``coalescence span''  greater than 0 and less than 100\%, indicating that there 
are nodes that are ``locally unary'' (Appendix~\ref{sec-locally-unary-edges}),
but mark a coalescence between lineages elsewhere along the sequence.

@@

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
Peter Ralph and Andrew Vaughn
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
We are particularly grateful to Graham Coop for detailed feedback
on and earlier version of the manuscript.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.

>>>>

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
% New developments have made it possible to infer ARGs at scale,
% enabling many new applications in population and statistical genetics.
% This rapid progress has led to a diversity of ARG definitions and representations.
Classical formalisms have focused on mapping 
coalescence and recombination events to the nodes in an ARG.
This approach is out of step with modern developments,
which do not represent genetic inheritance in terms of these events 
or explicitly infer them.
We present a simple formalism that defines an ARG in terms 
of specific genomes and their intervals of genetic inheritance,
and show how it generalises these classical treatments and 
encompasses the outputs of recent methods. We discuss 
nuances arising from this more general structure, and argue that it
forms an appropriate basis for a software standard in this rapidly 
growing field.
\end{abstract}

@@

\section{Event ARGs}
\label{sec-eARG}
A classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This Event ARG (eARG) encoding is the basis of the output formats 
created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG.
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
pastwards. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
we have three genome segments that are ancestral to the
four samples. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

\section{A diversity of structures}
\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
The ARGs in Fig.~\ref{fig-inferred-args}A and B are precise 
estimates (Appendix~\ref{sec-precision}),
with each node corresponding to a common ancestor 
or recombination event, or  equivalently, either having two children
or two parents.
In contrast the ARGs in Fig.~\ref{fig-inferred-args}C and D do not 
have this clear-cut interpretation, and the nodes
can simultaneously have more that than two children and more
than two parents. Another dimension of variability among the ARGs
is that the first three methods infer nodes that have
a ``coalescence span''  greater than 0 and less than 100\%, indicating that there 
are nodes that are ``locally unary'' (Appendix~\ref{sec-locally-unary-edges}),
but mark a coalescence between lineages elsewhere along the sequence.

@@

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
Peter Ralph and Andrew Vaughn
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
We are particularly grateful to Graham Coop for detailed feedback
on an earlier version of the manuscript.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
